Component response rate variation underlies the stability of complex systems

### 2020-03-02 notebook/ms.tex
Edit the notation a bit in the Introduction to make the results easier to read. Also, add some more analytical explanation to the start of the results

\textbf{Component response rates of random complex systems}. Complex
systems \(\mathbf{M}\) are built from two matrices, one modelling
component interactions (\(\mathbf{A}\)), and second modelling component
response rates (\(\gamma\)). Both \(\mathbf{A}\) and \(\gamma\) are
square \(S \times S\) matrices. Rows in \(\mathbf{A}\) define how a
given component \(i\) is affected by each component \(j\) in the system,
including itself (where \(i = j\)). Off-diagonal elments of
\(\mathbf{A}\) are independent and identically distributed (i.i.d), and
diagonal elements are set to \(A_{ii} = -1\) as in
May\textsuperscript{\protect\hyperlink{ref-May1972}{1}}. Diagonal
elements of \(\gamma\) are positive, and off-diagonal elements are set
to zero (i.e. \(\gamma\) is a diagonal matrix with positive support).
The distribution of \(\gamma\) over \(S\) components thereby models the
distribution of component response rates. The dynamics of the entire
system \(\mathbf{M}\) can be defined
as\textsuperscript{\protect\hyperlink{ref-Patel2018}{18}},

@@

Equation \ref{defM} thereby serves as a null model to investigate how
varying component response rates affect complex systems.

Higher stability in systems with variation in \(\gamma\) can be observed
by sampling \(\gamma_{i}\) values from various distributions. I focus on
a uniform distribution where \(\gamma \sim \mathcal{U}(0, 2)\) (see
Supplementary Information for other distributions, which give similar
results). As with the case of \(\gamma = \{1.95, 0.5\}\) (Fig. 1b), mean
\(\gamma = 1\) when \(\gamma \sim \mathcal{U}(0, 2)\), allowing
comparison of \(\mathbf{M}\) before and after the addition of variation
in component response rate. Figure 2 shows a comparison of eigenvalue
distributions given \(S = 1000\), \(C = 0.05\), and \(\sigma = 0.4\). As
expected\textsuperscript{\protect\hyperlink{ref-Tao2010}{14}}, when
\(\gamma = 1\), eigenvalues are distributed uniformly in a circle
centred at (\(-1, 0\)) with a radius of \(\sigma\sqrt{SC} =\) 2.828.
Uniform variation in \(\gamma\) leads to a non-uniform distribution of
eigenvalues\textsuperscript{\protect\hyperlink{ref-Ahmadian2015}{19}--\protect\hyperlink{ref-Stone2017}{21}},
some of which are clustered locally near the centre of the distribution,
but others of which are spread outside the former radius of 2.828 (Fig
2b). The clustering and spreading of eigenvalues introduced by
\(Var(\gamma)\) can destabilise previously stable systems or stabilise
systems that are otherwise unstable. But where systems are otherwise too
complex to be stable given \(\gamma = 1\), the effect of \(Var(\gamma)\)
can often lead to stability above
May's\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2}}
threshold \(\sigma\sqrt{SC} < 1\).

@@

\textbf{Targeted manipulation of \(\mathbf{\gamma}\)}. To further
investigate the potential of \(Var(\gamma)\) to be stabilising, I used a
genetic algorithm. Genetic algorithms are heuristic tools that mimic
evolution by natural selection, and are useful when the space of
potential solutions (in this case, possible combinations of \(\gamma\)
values leading to stability in a complex system) is too large to search
exhaustively\textsuperscript{\protect\hyperlink{ref-Hamblin2013}{22}}.
Generations of selection on \(\gamma\) value combinations to minimise
\(\max\left(\Re(\lambda)\right)\) demonstrated the potential for
\(Var(\gamma)\) to increase system stability. Across
\(S = \{2, 3, ..., 39, 40\}\), sets of \(\gamma\) values were found that
resulted in stable systems with probabilities that were up to four
orders of magnitude higher than when \(\gamma = 1\) (Fig. 4), meaning
that stability could often be achieved by manipulating \(S\) \(\gamma\)
values rather than \(S \times S\) \(\mathbf{M}\) elements (i.e., by
manipulating component response rates rather than interactions between
components).

\textbf{System feasibility given \(\mathbf{Var(\gamma)}\)} For complex
systems in which individual system components represent the density of
some tangible quantity, it is relevant to consider the feasibility of
the system. Feasibilility assumes that values of all components are
positive at
equilibrium\textsuperscript{\protect\hyperlink{ref-Grilli2017}{5},\protect\hyperlink{ref-Dougoud2018}{23},\protect\hyperlink{ref-Song2018}{24}}.
This is of particular interest for ecological communities because
population density (\(N\)) cannot take negative values, meaning that
ecological systems need to be feasible for stability to be biologically
realistic\textsuperscript{\protect\hyperlink{ref-Dougoud2018}{23}}.
While my results are intended to be general to all complex systems, and
not restricted to species networks, I have also performed a feasibility
analysis on all matrices \(\mathbf{M}\) tested for stability, and
additionally for specific types of ecological
communities\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2}}
(e.g., competitive, mutualist, predator-prey; see Supplementary
Information). I emphasise that \(\gamma\) is not interpreted as
population density in this analysis, but instead as a fundamental
property of species life history such as expected generation time.
Feasibility was unaffected by \(Var(\gamma)\) and instead occurred with
a fixed probability of \(1/2^{S}\), consistent with a recent proof by
Serván et al.\textsuperscript{\protect\hyperlink{ref-Servan2018}{25}}
(see Supplementary Information). Hence, for pure interacting species
networks, variation in component response rate (i.e., species generation
time) does not affect stability at biologically realistic species
densities.

@@

My focus is distinct from Gibbs et
al.\textsuperscript{\protect\hyperlink{ref-Gibbs2017}{20}}, who applied
the same mathematical framework to investigate how a diagonal matrix
\(\mathbf{X}\) (equivalent to \(\gamma\) in my model) affects the
stability of a community matrix \(\mathbf{M}\) given an interaction
matrix \(\mathbf{A}\) within a generalised Lotka-Volterra model, where
\(\mathbf{M} = \mathbf{XA}\). Gibbs et
al.\textsuperscript{\protect\hyperlink{ref-Gibbs2017}{20}} analytically
demonstrated that the effect of \(\mathbf{X}\) on system stability
decreases exponentially as system size becomes arbitrarily large
(\(S \to \infty\)) for a given magnitude of complexity
\(\sigma\sqrt{SC}\). My numerical results do not contradict this
prediction because I did not scale \(\sigma = 1 / \sqrt{S}\), but
instead fixed \(\sigma\) and increased \(S\) to thereby increase total
system complexity (see Supplemental Information for results simulated
across \(\sigma\) and \(C\)). Overall, I show that component response
rate variation increases the upper bound of complexity at which
stability can be realistically observed, meaning that highly complex
systems are more likely than not to vary in their component response
rates, and for this variation to be critical for system stability.

@@

Interestingly, while complex systems were more likely to be stable given
variation in component response rate, they were not more likely to be
feasible, meaning that stability was not increased when component values
were also restricted to being positive at equilibrium. Feasibility is
important to consider, particularly for the study of ecological networks
of
species\textsuperscript{\protect\hyperlink{ref-Grilli2017}{5},\protect\hyperlink{ref-Stone2017}{21},\protect\hyperlink{ref-Dougoud2018}{23},\protect\hyperlink{ref-Servan2018}{25}}
because population densities cannot realistically be negative. My
results therefore suggest that variation in the rate of population
responses to perturbation (e.g., due to differences in generation time
among species) is unlikely to be critical to the stability of purely
multi-species interaction networks (see also Supplementary Information).
Nevertheless, ecological interactions do not exist in isolation in
empirical
systems\textsuperscript{\protect\hyperlink{ref-Patel2018}{18}}, but
instead interact with evolutionary, abiotic, or social-economic systems.
The relevance of component response rate for complex system stability
should therefore not be ignored in the broader context of ecological
communities.

A general mathematical framework encompassing shifts in eigenvalue
distributions caused by a vector \(\gamma\) has been
investigated\textsuperscript{\protect\hyperlink{ref-Ahmadian2015}{19}}
and recently applied to questions concerning species density and
feasibility\textsuperscript{\protect\hyperlink{ref-Gibbs2017}{20},\protect\hyperlink{ref-Stone2017}{21}},
but \(\gamma\) has not been interpreted as rates of response of
individual system components to perturbation. My model focuses on
component response rates for systems of a finite size, in which
complexity is high but not yet high enough to make the probability of
stability unrealistically low for actual empirical systems. For this
upper range of system size, randomly assembled complex systems are more
likely to be stable if their component response rates vary (e.g.,
\(10 < S < 30\) for parameter values in Fig. 3). Overall, I suggest that
variation in component response rate might therefore be critical for
maintaining stability in many highly complex empirical systems. These
results are broadly applicable for understanding the stability of
complex networks across the physical, life, and social sciences.

@@

I use the same principle that Patel et al. use to modulate the relative
rate of evolution to modulate rates of component responses for \(S\)
components. Following
May\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-May1973}{26}},
the value of a component \(i\) at time \(t\) (\(v_{i}(t)\)) is affected
by the value of \(j\) (\(v_{j}(t)\)) and \(j\)'s marginal effect on
\(i\) (\(a_{ij}\)), and by \(i\)'s response rate (\(\gamma_{i}\)),

@@

In matrix notation\textsuperscript{\protect\hyperlink{ref-May1973}{26}},

@@

In the above, \(\mathbf{\gamma}\) is a diagonal matrix in which elements
correspond to individual component response rates. Therefore,
\(\mathbf{M} = \mathbf{\gamma} \mathbf{A}\) defines the values of system
components and can be analysed using the techniques of
May\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Ahmadian2015}{19},\protect\hyperlink{ref-May1973}{26}}.
In these analyses, row means of \(\mathbf{A}\) are expected to be
identical, but variation around this expectation will naturally arise
due to random sampling of \(\mathbf{A}\) off-diagonal elements and
finite \(S\). In simulations, the total variation in \(\mathbf{M}\) row
means that is attributable to \(\mathbf{A}\) is small relative to that
attributable to \(\mathbf{\gamma}\), especially at high \(S\). Variation
in \(\mathbf{\gamma}\) specifically isolates the effects of differing
component response rates, hence causing differences in expected
\(\mathbf{M}\) row means.

\textbf{Genetic algorithm}. Ideally, to investigate the potential of
\(Var(\gamma)\) for increasing the proportion of stable complex systems,
the search space of all possible \(\gamma\) vectors would be evaluated
for each unique \(\mathbf{M = \gamma A}\). This is technically
impossible because \(\gamma_{i}\) can take any real value between 0-2,
but even rounding \(\gamma_{i}\) to reasonable values would result in a
search space too large to practically explore. Under these conditions,
genetic algorithms are highly useful tools for finding practical
solutions by mimicking the process of biological
evolution\textsuperscript{\protect\hyperlink{ref-Hamblin2013}{22}}. In
this case, the practical solution is finding vectors of
\(\mathbf{\gamma}\) that decrease the most positive real eigenvalue of
\(\mathbf{M}\). The genetic algorithm used achieves this by initialising
a large population of 1000 different potential \(\mathbf{\gamma}\)
vectors and allowing this population to evolve through a process of
mutation, crossover (swaping \(\gamma_{i}\) values between vectors),
selection, and reproduction until either a \(\mathbf{\gamma}\) vector is
found where all \(\Re(\lambda) < 0\) or some ``giving up'' critiera is
met.

@@

\textbf{System feasibility}. Dougoud et
al.\textsuperscript{\protect\hyperlink{ref-Dougoud2018}{23}} identify
the following feasibility criteria for ecological systems characterised
by \(S\) interacting species with varying densities in a generalised
Lotka-Volterra model,

@@

In the above, \(\mathbf{n^{*}}\) is the vector of species densities at
equilibrium. Feasibility is satisfied if all elements in
\(\mathbf{n^{*}}\) are positive. The matrix \(\mathbf{I}\) is the
identity matrix, and the value \(\theta\) is the strength of
intraspecific competition (diagonal elements). Diagonal values are set
to \(-1\), so \(\theta = -1\). The variable \(\delta\) is a
normalisation parameter that modulates the strength of interactions
(\(\sigma\)) for \(\mathbf{J}\). Implicitly, here \(\delta = 0\)
underlying strong interactions. Hence, \((CS)^{-\delta} = 1\), so in the
above, a diagonal matrix of -1s (\(\theta \mathbf{I}\)) is added to
\(\mathbf{J}\), which has a diagonal of all zeros and an off-diagonal
affecting species interactions (i.e., the expression \((CS)^{-\delta}\)
relates to May's\textsuperscript{\protect\hyperlink{ref-May1972}{1}}
stability
criterion\textsuperscript{\protect\hyperlink{ref-Dougoud2018}{23}} by
\(\frac{\sigma}{(CS)^{-\delta}}\sqrt{SC} < 1\), and hence for my
purposes \((CS)^{-\delta} = 1\)). Given
\(\mathbf{A} = \theta\mathbf{I + J}\), the above criteria is therefore
reduced to the below (see
also\textsuperscript{\protect\hyperlink{ref-Servan2018}{25}}),

@@

\hypertarget{ref-Ahmadian2015}{}
19. Ahmadian, Y., Fumarola, F. \& Miller, K. D. Properties of networks
with partially structured and partially random connectivity.
\emph{Physical Review E - Statistical, Nonlinear, and Soft Matter
Physics} \textbf{91,} 012820 (2015).

\hypertarget{ref-Gibbs2017}{}
20. Gibbs, T., Grilli, J., Rogers, T. \& Allesina, S. The effect of
population abundances on the stability of large random ecosystems.
\emph{Physical Review E - Statistical, Nonlinear, and Soft Matter
Physics} \textbf{98,} 022410 (2018).

\hypertarget{ref-Stone2017}{}
21. Stone, L. The feasibility and stability of large complex biological
networks: a random matrix approach. \emph{Scientific Reports}
\textbf{8,} 8246 (2018).

\hypertarget{ref-Hamblin2013}{}
22. Hamblin, S. On the practical usage of genetic algorithms in ecology
and evolution. \emph{Methods in Ecology and Evolution} \textbf{4,}
184--194 (2013).

\hypertarget{ref-Dougoud2018}{}
23. Dougoud, M., Vinckenbosch, L., Rohr, R., Bersier, L.-F. \& Mazza, C.
The feasibility of equilibria in large ecosystems: a primary but
neglected concept in the complexity-stability debate. \emph{PLOS
Computational Biology} \textbf{14,} e1005988 (2018).

\hypertarget{ref-Song2018}{}
24. Song, C. \& Saavedra, S. Will a small randomly assembled community
be feasible and stable? \emph{Ecology} \textbf{99,} 743--751 (2018).

\hypertarget{ref-Servan2018}{}
25. Serván, C. A., Capitán, J. A., Grilli, J., Morrison, K. E. \&
Allesina, S. Coexistence of many species in random ecosystems.
\emph{Nature Ecology and Evolution} \textbf{2,} 1237--1242 (2018).

\hypertarget{ref-May1973}{}
26. May, R. M. Qualitative stability in model ecosystems. \emph{Ecology}
\textbf{54,} 638--641 (1973).

@@

\textbf{Figure 2: Distributions of eigenvalues before (a) and after (b)
introducing variation in component response rate
(\(\boldsymbol{\gamma}\)) in complex systems.} Each panel show the same
system where \(S = 1000\), \(C = 0.05\), and \(\sigma = 0.4\).
\textbf{a.} Eigenvalues plotted in the absence of \(Var(\gamma)\) where
\(E[\gamma] = 1\), versus \textbf{b.} eigenvalues plotted given
\(\gamma \sim \mathcal{U}(0, 2)\), which increases the variance of
interaction strengths (\(\sigma^{2}\)) but also creates a cluster of
eigenvalues toward the distribution's centre (-1, 0). Blue elipses in
both panels show the circle centred on the distribution in panel a.
Proportions of \(\Re(\lambda) < 0\) are 0.724 and 0.738 for a and b,
respectively.

>>>>

\textbf{Component response rates of random complex systems}. Complex
systems \(\mathbf{M}\) are built from two matrices, one modelling
component interactions (\(\mathbf{A}\)), and second modelling component
response rates (\(\gamma\)). Both \(\mathbf{A}\) and \(\gamma\) are
square \(S \times S\) matrices. Rows in \(\mathbf{A}\) define how a
given component \(i\) is affected by each component \(j\) in the system,
including itself (where \(i = j\)). Off-diagonal elments of
\(\mathbf{A}\) are independent and identically distributed (i.i.d), and
diagonal elements are set to \(A_{ii} = -1\) as in
May\textsuperscript{\protect\hyperlink{ref-May1972}{1}}. Diagonal
elements of \(\gamma\) are positive, and off-diagonal elements are set
to zero (i.e. \(\gamma\) is a diagonal matrix with positive support).
The distribution of \(\gamma\) over \(S\) components thereby models the
distribution of component response rates. The dynamics of the entire
system \(\mathbf{M}\) can be defined as
follows\textsuperscript{\protect\hyperlink{ref-Patel2018}{18}},

@@

Equation \ref{defM} thereby serves as a null model to investigate how
variation in component response rate (\(\sigma_{\gamma}\)) affect
complex systems. In the absence of such variation
(\(\sigma_{\gamma} = 0\)), \(\gamma\) is set to the identity matrix
(diagonal elements all equal 1) and \(\textbf{M} = \textbf{A}\). Given
this baseline condition, \(\max(\Re(\lambda))\) for
\(\textbf{M} = \textbf{A}\) can be estimated from five
values\textsuperscript{\protect\hyperlink{ref-Tang2014b}{19}}. These
values include (1) system size (\(S\)), (2) mean self-regulation of
components (\(d\); i.e., mean diagonal element of \(\textbf{M}\)), (3)
mean interaction strength between components (\(\mu\)), (4) the standard
deviation between component interaction strengths (\(\sigma\)), and (5)
the correlation of interaction strengths between components, \(M_{ij}\)
and \(M_{ji}\) (\(\rho\)).

When investigating the effect of varying component response rate
\(\sigma_{\gamma}\) on stability by defining
\(\textbf{M} = \gamma\textbf{A}\), \(S\) remains unchanged. Further,
values of \(\gamma_{i}\) were sampled such that \(E[d]\) and \(E[\mu]\)
also remained unchanged (in practice, diagonal elements of
\(\textbf{M}\) were standardised so that mean values were identical
before and after adding \(\gamma\)). What \(\sigma_{\gamma}\) does
change is the variation in component interaction strengths, and
\(\rho\). Introducing variation in \(\gamma\) increases the total
variation in the system, making it more likely that
\(\textbf{M} = \gamma\textbf{A}\) is unstable. Variation of the
off-diagonal elements in \(\textbf{M}\) is described by the joint
variation of two random variables (to simplify the notation,
\(\sigma^{2}_{M}\) and \(\sigma^{2}_{A}\) refer to variances of
off-diagonal elements only),

\begin{equation} \label{var_full}
\sigma^{2}_{M} = \sigma^{2}_{A}\sigma^{2}_{\gamma} + \sigma^{2}_{A}E[\gamma_{i}]^{2}+\sigma^{2}_{\gamma}E[A_{ij}]^{2}.
\end{equation}

Note that in my simulations \(E[\gamma_{i}] = 1\) and \(E[A_{ij}] = 0\),
so equation \ref{var_full} can be simplified,

\begin{equation} \label{var_reduced}
\sigma^{2}_{M} = \sigma^{2}_{A}(1 + \sigma^{2}_{\gamma}).
\end{equation}

The increase caused by \(\sigma^{2}_\gamma\) can be visualised from the
eigenvalue spectra of \(\textbf{A}\) versus
\(\textbf{M} = \gamma\textbf{A}\). Given \(d = 0\) and \(C = 1\), the
distribution of eigenvalues of \(\textbf{M}\) and \(\textbf{A}\) lie
within a circle of a radius \(\sigma^{2}_{M}\sqrt{S}\) and
\(\sigma^{2}_{A}\sqrt{S}\), respectively.

Higher stability in systems with variation in \(\gamma\) can be observed
by sampling \(\gamma_{i}\) values from various distributions. I focus on
a uniform distribution where \(\gamma \sim \mathcal{U}(0, 2)\) (see
Supplementary Information for other distributions, which give similar
results). As with the case of \(\gamma = \{1.95, 0.5\}\) (Fig. 1b), mean
\(\gamma = 1\) when \(\gamma \sim \mathcal{U}(0, 2)\), allowing
comparison of \(\mathbf{M}\) before and after the addition of variation
in component response rate. Figure 2 shows a comparison of eigenvalue
distributions given \(S = 1000\), \(C = 0.05\), and \(\sigma = 0.4\). As
expected\textsuperscript{\protect\hyperlink{ref-Tao2010}{14}}, when
\(\gamma = 1\), eigenvalues are distributed uniformly in a circle
centred at (\(-1, 0\)) with a radius of \(\sigma\sqrt{SC} =\) 2.828.
Uniform variation in \(\gamma\) leads to a non-uniform distribution of
eigenvalues\textsuperscript{\protect\hyperlink{ref-Ahmadian2015}{20}--\protect\hyperlink{ref-Stone2017}{22}},
some of which are clustered locally near the centre of the distribution,
but others of which are spread outside the former radius of 2.828 (Fig
2b). The clustering and spreading of eigenvalues introduced by
\(Var(\gamma)\) can destabilise previously stable systems or stabilise
systems that are otherwise unstable. But where systems are otherwise too
complex to be stable given \(\gamma = 1\), the effect of \(Var(\gamma)\)
can often lead to stability above
May's\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2}}
threshold \(\sigma\sqrt{SC} < 1\).

@@

\textbf{Targeted manipulation of \(\mathbf{\gamma}\)}. To further
investigate the potential of \(Var(\gamma)\) to be stabilising, I used a
genetic algorithm. Genetic algorithms are heuristic tools that mimic
evolution by natural selection, and are useful when the space of
potential solutions (in this case, possible combinations of \(\gamma\)
values leading to stability in a complex system) is too large to search
exhaustively\textsuperscript{\protect\hyperlink{ref-Hamblin2013}{23}}.
Generations of selection on \(\gamma\) value combinations to minimise
\(\max\left(\Re(\lambda)\right)\) demonstrated the potential for
\(Var(\gamma)\) to increase system stability. Across
\(S = \{2, 3, ..., 39, 40\}\), sets of \(\gamma\) values were found that
resulted in stable systems with probabilities that were up to four
orders of magnitude higher than when \(\gamma = 1\) (Fig. 4), meaning
that stability could often be achieved by manipulating \(S\) \(\gamma\)
values rather than \(S \times S\) \(\mathbf{M}\) elements (i.e., by
manipulating component response rates rather than interactions between
components).

\textbf{System feasibility given \(\mathbf{Var(\gamma)}\)} For complex
systems in which individual system components represent the density of
some tangible quantity, it is relevant to consider the feasibility of
the system. Feasibilility assumes that values of all components are
positive at
equilibrium\textsuperscript{\protect\hyperlink{ref-Grilli2017}{5},\protect\hyperlink{ref-Dougoud2018}{24},\protect\hyperlink{ref-Song2018}{25}}.
This is of particular interest for ecological communities because
population density (\(N\)) cannot take negative values, meaning that
ecological systems need to be feasible for stability to be biologically
realistic\textsuperscript{\protect\hyperlink{ref-Dougoud2018}{24}}.
While my results are intended to be general to all complex systems, and
not restricted to species networks, I have also performed a feasibility
analysis on all matrices \(\mathbf{M}\) tested for stability, and
additionally for specific types of ecological
communities\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2}}
(e.g., competitive, mutualist, predator-prey; see Supplementary
Information). I emphasise that \(\gamma\) is not interpreted as
population density in this analysis, but instead as a fundamental
property of species life history such as expected generation time.
Feasibility was unaffected by \(Var(\gamma)\) and instead occurred with
a fixed probability of \(1/2^{S}\), consistent with a recent proof by
Serván et al.\textsuperscript{\protect\hyperlink{ref-Servan2018}{26}}
(see Supplementary Information). Hence, for pure interacting species
networks, variation in component response rate (i.e., species generation
time) does not affect stability at biologically realistic species
densities.

@@

My focus is distinct from Gibbs et
al.\textsuperscript{\protect\hyperlink{ref-Gibbs2017}{21}}, who applied
the same mathematical framework to investigate how a diagonal matrix
\(\mathbf{X}\) (equivalent to \(\gamma\) in my model) affects the
stability of a community matrix \(\mathbf{M}\) given an interaction
matrix \(\mathbf{A}\) within a generalised Lotka-Volterra model, where
\(\mathbf{M} = \mathbf{XA}\). Gibbs et
al.\textsuperscript{\protect\hyperlink{ref-Gibbs2017}{21}} analytically
demonstrated that the effect of \(\mathbf{X}\) on system stability
decreases exponentially as system size becomes arbitrarily large
(\(S \to \infty\)) for a given magnitude of complexity
\(\sigma\sqrt{SC}\). My numerical results do not contradict this
prediction because I did not scale \(\sigma = 1 / \sqrt{S}\), but
instead fixed \(\sigma\) and increased \(S\) to thereby increase total
system complexity (see Supplemental Information for results simulated
across \(\sigma\) and \(C\)). Overall, I show that component response
rate variation increases the upper bound of complexity at which
stability can be realistically observed, meaning that highly complex
systems are more likely than not to vary in their component response
rates, and for this variation to be critical for system stability.

@@

Interestingly, while complex systems were more likely to be stable given
variation in component response rate, they were not more likely to be
feasible, meaning that stability was not increased when component values
were also restricted to being positive at equilibrium. Feasibility is
important to consider, particularly for the study of ecological networks
of
species\textsuperscript{\protect\hyperlink{ref-Grilli2017}{5},\protect\hyperlink{ref-Stone2017}{22},\protect\hyperlink{ref-Dougoud2018}{24},\protect\hyperlink{ref-Servan2018}{26}}
because population densities cannot realistically be negative. My
results therefore suggest that variation in the rate of population
responses to perturbation (e.g., due to differences in generation time
among species) is unlikely to be critical to the stability of purely
multi-species interaction networks (see also Supplementary Information).
Nevertheless, ecological interactions do not exist in isolation in
empirical
systems\textsuperscript{\protect\hyperlink{ref-Patel2018}{18}}, but
instead interact with evolutionary, abiotic, or social-economic systems.
The relevance of component response rate for complex system stability
should therefore not be ignored in the broader context of ecological
communities.

A general mathematical framework encompassing shifts in eigenvalue
distributions caused by a vector \(\gamma\) has been
investigated\textsuperscript{\protect\hyperlink{ref-Ahmadian2015}{20}}
and recently applied to questions concerning species density and
feasibility\textsuperscript{\protect\hyperlink{ref-Gibbs2017}{21},\protect\hyperlink{ref-Stone2017}{22}},
but \(\gamma\) has not been interpreted as rates of response of
individual system components to perturbation. My model focuses on
component response rates for systems of a finite size, in which
complexity is high but not yet high enough to make the probability of
stability unrealistically low for actual empirical systems. For this
upper range of system size, randomly assembled complex systems are more
likely to be stable if their component response rates vary (e.g.,
\(10 < S < 30\) for parameter values in Fig. 3). Overall, I suggest that
variation in component response rate might therefore be critical for
maintaining stability in many highly complex empirical systems. These
results are broadly applicable for understanding the stability of
complex networks across the physical, life, and social sciences.

@@

I use the same principle that Patel et al. use to modulate the relative
rate of evolution to modulate rates of component responses for \(S\)
components. Following
May\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-May1973}{27}},
the value of a component \(i\) at time \(t\) (\(v_{i}(t)\)) is affected
by the value of \(j\) (\(v_{j}(t)\)) and \(j\)'s marginal effect on
\(i\) (\(a_{ij}\)), and by \(i\)'s response rate (\(\gamma_{i}\)),

@@

In matrix notation\textsuperscript{\protect\hyperlink{ref-May1973}{27}},

@@

In the above, \(\mathbf{\gamma}\) is a diagonal matrix in which elements
correspond to individual component response rates. Therefore,
\(\mathbf{M} = \mathbf{\gamma} \mathbf{A}\) defines the values of system
components and can be analysed using the techniques of
May\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Ahmadian2015}{20},\protect\hyperlink{ref-May1973}{27}}.
In these analyses, row means of \(\mathbf{A}\) are expected to be
identical, but variation around this expectation will naturally arise
due to random sampling of \(\mathbf{A}\) off-diagonal elements and
finite \(S\). In simulations, the total variation in \(\mathbf{M}\) row
means that is attributable to \(\mathbf{A}\) is small relative to that
attributable to \(\mathbf{\gamma}\), especially at high \(S\). Variation
in \(\mathbf{\gamma}\) specifically isolates the effects of differing
component response rates, hence causing differences in expected
\(\mathbf{M}\) row means.

\textbf{Genetic algorithm}. Ideally, to investigate the potential of
\(Var(\gamma)\) for increasing the proportion of stable complex systems,
the search space of all possible \(\gamma\) vectors would be evaluated
for each unique \(\mathbf{M = \gamma A}\). This is technically
impossible because \(\gamma_{i}\) can take any real value between 0-2,
but even rounding \(\gamma_{i}\) to reasonable values would result in a
search space too large to practically explore. Under these conditions,
genetic algorithms are highly useful tools for finding practical
solutions by mimicking the process of biological
evolution\textsuperscript{\protect\hyperlink{ref-Hamblin2013}{23}}. In
this case, the practical solution is finding vectors of
\(\mathbf{\gamma}\) that decrease the most positive real eigenvalue of
\(\mathbf{M}\). The genetic algorithm used achieves this by initialising
a large population of 1000 different potential \(\mathbf{\gamma}\)
vectors and allowing this population to evolve through a process of
mutation, crossover (swaping \(\gamma_{i}\) values between vectors),
selection, and reproduction until either a \(\mathbf{\gamma}\) vector is
found where all \(\Re(\lambda) < 0\) or some ``giving up'' critiera is
met.

@@

\textbf{System feasibility}. Dougoud et
al.\textsuperscript{\protect\hyperlink{ref-Dougoud2018}{24}} identify
the following feasibility criteria for ecological systems characterised
by \(S\) interacting species with varying densities in a generalised
Lotka-Volterra model,

@@

In the above, \(\mathbf{n^{*}}\) is the vector of species densities at
equilibrium. Feasibility is satisfied if all elements in
\(\mathbf{n^{*}}\) are positive. The matrix \(\mathbf{I}\) is the
identity matrix, and the value \(\theta\) is the strength of
intraspecific competition (diagonal elements). Diagonal values are set
to \(-1\), so \(\theta = -1\). The variable \(\delta\) is a
normalisation parameter that modulates the strength of interactions
(\(\sigma\)) for \(\mathbf{J}\). Implicitly, here \(\delta = 0\)
underlying strong interactions. Hence, \((CS)^{-\delta} = 1\), so in the
above, a diagonal matrix of -1s (\(\theta \mathbf{I}\)) is added to
\(\mathbf{J}\), which has a diagonal of all zeros and an off-diagonal
affecting species interactions (i.e., the expression \((CS)^{-\delta}\)
relates to May's\textsuperscript{\protect\hyperlink{ref-May1972}{1}}
stability
criterion\textsuperscript{\protect\hyperlink{ref-Dougoud2018}{24}} by
\(\frac{\sigma}{(CS)^{-\delta}}\sqrt{SC} < 1\), and hence for my
purposes \((CS)^{-\delta} = 1\)). Given
\(\mathbf{A} = \theta\mathbf{I + J}\), the above criteria is therefore
reduced to the below (see
also\textsuperscript{\protect\hyperlink{ref-Servan2018}{26}}),

@@

\hypertarget{ref-Tang2014b}{}
19. Tang, S. \& Allesina, S. Reactivity and stability of large
ecosystems. \emph{Frontiers in Ecology and Evolution} \textbf{2,} 1--8
(2014).

\hypertarget{ref-Ahmadian2015}{}
20. Ahmadian, Y., Fumarola, F. \& Miller, K. D. Properties of networks
with partially structured and partially random connectivity.
\emph{Physical Review E - Statistical, Nonlinear, and Soft Matter
Physics} \textbf{91,} 012820 (2015).

\hypertarget{ref-Gibbs2017}{}
21. Gibbs, T., Grilli, J., Rogers, T. \& Allesina, S. The effect of
population abundances on the stability of large random ecosystems.
\emph{Physical Review E - Statistical, Nonlinear, and Soft Matter
Physics} \textbf{98,} 022410 (2018).

\hypertarget{ref-Stone2017}{}
22. Stone, L. The feasibility and stability of large complex biological
networks: a random matrix approach. \emph{Scientific Reports}
\textbf{8,} 8246 (2018).

\hypertarget{ref-Hamblin2013}{}
23. Hamblin, S. On the practical usage of genetic algorithms in ecology
and evolution. \emph{Methods in Ecology and Evolution} \textbf{4,}
184--194 (2013).

\hypertarget{ref-Dougoud2018}{}
24. Dougoud, M., Vinckenbosch, L., Rohr, R., Bersier, L.-F. \& Mazza, C.
The feasibility of equilibria in large ecosystems: a primary but
neglected concept in the complexity-stability debate. \emph{PLOS
Computational Biology} \textbf{14,} e1005988 (2018).

\hypertarget{ref-Song2018}{}
25. Song, C. \& Saavedra, S. Will a small randomly assembled community
be feasible and stable? \emph{Ecology} \textbf{99,} 743--751 (2018).

\hypertarget{ref-Servan2018}{}
26. Serván, C. A., Capitán, J. A., Grilli, J., Morrison, K. E. \&
Allesina, S. Coexistence of many species in random ecosystems.
\emph{Nature Ecology and Evolution} \textbf{2,} 1237--1242 (2018).

\hypertarget{ref-May1973}{}
27. May, R. M. Qualitative stability in model ecosystems. \emph{Ecology}
\textbf{54,} 638--641 (1973).

@@

\textbf{Figure 2: Distributions of eigenvalues before (a) and after (b)
introducing variation in component response rate
(\(\boldsymbol{\gamma}\)) in complex systems.} Each panel show the same
system where \(S = 1000\), \(C = 0.05\), and \(\sigma = 0.4\).
\textbf{a.} Eigenvalues plotted in the absence of \(Var(\gamma)\) where
\(E[\gamma] = 1\), versus \textbf{b.} eigenvalues plotted given
\(\gamma \sim \mathcal{U}(0, 2)\), which increases the variance of
interaction strengths (\(\sigma^{2}\)) but also creates a cluster of
eigenvalues toward the distribution's centre (-1, 0). Blue elipses in
both panels show the circle centred on the distribution in panel a.
Proportions of \(\Re(\lambda) < 0\) are 0.718 and 0.742 for a and b,
respectively.
